A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-12-23 podc_paper/algorithm.tex
Added some diagrams

\subsection{Overview of Implementation}
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
In this section, we assume for simplicity that each node has an infinite blocks array.
Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.

@@

Now suppose $P$'s operation has been propagated all the way to the root.
If $P$'s operation is an enqueue, it has obtained a place in the linearization ordering and can terminate.
If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
(since the operation may have been propagated to the root by some other process).
Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
If so, it returns \nil\ and we call it a \emph{null dequeue}
If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
among all non-null dequeues,
finds the $r$th enqueue in the linearization, and returns that enqueue's value.

@@

@@

Finally, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
linearization ordering is 
$E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k)$.

\here{Need some figures illustrating tournament tree, ordering of operations, representation as blocks.  Might be able to modify Fig \ref{fig::set} and \ref{fig:block}.}

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]

\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]
\end{tikzpicture}
\caption[The operations merged in a \nf{Refresh} step with
  sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from
  left to right. In each internal node one can arbitrarily linearize
  the sets of concurrent operations  propagated together in a
  \nf{Refresh}. For example $op_4^1$ and $op_3^2$ have propagated
  together in one \nf{Propagate} step and they will be propagated up
  to the root together. Since their execution time intervals overlap,
  they can be linearized in any order.} 
\end{center}
\end{figure}

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture}
[level 1/.style={level distance=3.3cm,sibling distance=0cm},
	level 2/.style={level distance=2.5cm,sibling distance=0cm},
	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
\begin{small}

\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption[The operations merged in a \nf{Refresh} step with (left,
  right) blocks.]{\label{fig:block} Using blocks to represent
  operations. Blocks between two lines $||$ are propagated together to
  the parent. Each block consists of a pair (left, right) indicating
  the number of operations from the left and the right child,
  respectively. For example, (12,6) in the root contains  (10,2) from
  the left child and (6,0) from the right child. The third block in
  the root (8,36) is created by merging (5,3) from the left child and
  (14,6) and (0,16) from the right child. (5,3) is the superblock of
  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).} 
\end{figure}

\subsection{Designing a Block Representation to Solve Tasks \ref{construct} to \ref{findenqueue}}

@@

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block, which would be much too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how the fields of the blocks were chosen for this implicit representation. 
See Figure \ref{object-fields}.

@@

We now consider task \ref{findrank}.
First, we must determine whether the queue is empty when a dequeue occurs.
To facilitate this, each block in the root stores a \var{size} field that stores the number of elements
in the queue after all operations in the linearization ordering up to that block (inclusive) 
have been performed.
We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
$B_{d-1}.\var{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
as $B_{d-1}.\var{sum\sub{enq}}-B_{d-1}.\var{size}$.
Then, for any non-null dequeue in $B_d$, we can also determine its
rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
whose value it should return (among all enqueues).

@@

In an execution on a queue where no dequeue operation returns
\nf{null}, the $k$th dequeue returns the argument of the $k$th
enqueue. In the general case a dequeue returns \nf{null} if and only
if the queue size after the previous operation is 0. We refer to such
a dequeue as a \it{null dequeue}. If the dequeue is the $k$th non-null
dequeue, it returns the argument of the $k$th enqueue. Having the size
of the queue after an operation we can compute the number of non-null
dequeues from the number of enqueues before the operation. So, if we
store the size of the queue after each block of operations in the
root, we can compute the index of the enqueue whose argument is the
response to a given dequeue in constant time.

>>>>

\subsection{Overview of Implementation}
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{tournament} for an example.
In this section, we assume for simplicity that each node has an infinite blocks array.
Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.

@@

Now suppose $P$'s operation has been propagated all the way to the root.
If $P$'s operation is an enqueue, it has obtained a place in the linearization ordering and can terminate.
If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
(since the operation may have been propagated to the root by some other process).
Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
If so, it returns \var{null} and we call it a \emph{null dequeue}.
If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
among all non-null dequeues,
finds the $r$th enqueue in the linearization, and returns that enqueue's value.

@@

\begin{figure}[tb]
\input{tournamentTree.pdf_t}
\caption{An example tournament tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \var{blocks} arrays of the seven nodes.  The leftmost element of each \var{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still in progress.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
\end{figure}

\begin{figure}
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
The leaf blocks simply show the \var{element} field.
Internal blocks show the \var{sum\sub{enq}} and \var{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also contain the additional \var{size} field.
The \var{super} field is not shown.}
\end{figure}

@@

Finally, when designing a total order for the operations propagated to the root, we choose
to put each block's enqueues before its dequeues.
Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
linearization ordering is 
$E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k)$.

\subsection{Designing a Block Representation to Solve Tasks \ref{construct} to \ref{findenqueue}}
\label{sec:fields}

@@

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block, which would be much too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how the fields of the blocks were chosen for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} would actually be represented, and Figure \ref{object-fields} for the definitions of the fields of blocks and tree nodes.

@@

We now consider task \ref{findrank}.
First, we must determine whether the queue is empty when a dequeue occurs.
To facilitate this, each block in the root stores a \var{size} field that stores the number of elements
in the queue after all operations in the linearization ordering up to that block (inclusive) 
have been performed.
We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
$B_{d-1}.\var{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
as $B_{d-1}.\var{sum\sub{enq}}-B_{d-1}.\var{size}$.
Then, for any non-null dequeue in $B_d$, we can use this information to determine its
rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
whose value it should return (among all enqueues).

@@
